---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rc3helpers

<!-- badges: start -->
<!-- badges: end -->

rc3helpers is a little package designed around making it easier to take the output of a custom `recount-unify` run, and create the directory structure that
will be easily imported into the [recount3](https://bioconductor.org/packages/release/bioc/html/recount3.html) R package.

recount3 has a very, very specific directory structure it wants for importing data.
This directory is not trivial to generate after you've done a recount-unify.

Two important notes:

1. As of recount3 v 1.22.0, there is no option to have an organism outside of *human* or *mouse*, so we always use a directory of one of those two organisms.
1. This also means that only certain annotations / references are supported. Therefore, we create symbolic links to the files that `recount3` expects to be there.

## Installation

You can install the development version of `rc3helpers` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("MoseleyBioinformaticsLab/rc3helpers")
```

## Example

I previously ran two SRA samples in local mode through the pump and unify workflow, with the `short_id` **sratest**.

```{r}
#| label: setup-directories
library(rc3helpers)
tmp_dir = tempdir()
rc3_dirs = rc3_setup_directory(base_dir = tmp_dir, short_id = "sratest")
```

The directories created should look like this:

```
human
├── annotations
│   ├── exon_sums
│   └── gene_sums
└── data_sources
    └── sratest
        ├── base_sums
        ├── exon_sums
        ├── gene_sums
        ├── junctions
        └── metadata
```

The unify output looks like this:

```
├── exon_sums_per_study
│   └── st
│       └── sratest
│           ├── sratest.exon_sums.sratest.ERCC.gz
│           ├── sratest.exon_sums.sratest.F006.gz
│           ├── sratest.exon_sums.sratest.G026.gz
│           ├── sratest.exon_sums.sratest.G029.gz
│           ├── sratest.exon_sums.sratest.R109.gz
│           └── sratest.exon_sums.sratest.SIRV.gz
├── gene_sums_per_study
│   └── st
│       └── sratest
│           ├── sratest.gene_sums.sratest.ERCC.gz
│           ├── sratest.gene_sums.sratest.F006.gz
│           ├── sratest.gene_sums.sratest.G026.gz
│           ├── sratest.gene_sums.sratest.G029.gz
│           ├── sratest.gene_sums.sratest.R109.gz
│           └── sratest.gene_sums.sratest.SIRV.gz
├── ids.tsv
├── junction_counts_per_study
│   └── st
│       └── sratest
│           ├── sratest.junctions.sratest.ALL.ID.gz
│           ├── sratest.junctions.sratest.ALL.MM.gz
│           ├── sratest.junctions.sratest.ALL.RR.gz
│           ├── sratest.junctions.sratest.UNIQUE.ID.gz
│           ├── sratest.junctions.sratest.UNIQUE.MM.gz
│           └── sratest.junctions.sratest.UNIQUE.RR.gz
├── junctions.bgz
├── junctions.bgz.tbi
├── junctions.sqlite
|
|
|
```

Where this is a model organism, you are going to need the reference sets of things downloaded too.
We can download the gene and exon sum annotations from recount3, in our case for G026.

```
wget http://duffel.rail.bio/recount3/human/new_annotations/gene_sums/human.gene_sums.G026.gtf.gz
wget http://duffel.rail.bio/recount3/human/new_annotations/exon_sums/human.exon_sums.G026.gtf.gz
```

We can pass the list of directories created by `rc3_setup_directories` to `rc3_copy_unify_outputs`:

```{r}
#| label: copy-unify
#| eval: false
rc3_dirs |>
  rc3_copy_unify_outputs(
    short_id = "sratest",
    unify_directory = "/path/to/unify",
    references_directory = "/path/to/references/hg38"
  )
```

Which will create this directory structure. You may see an error when working with human data, as the function tries to link a file that actually already exists.
That's OK.

```
human
├── annotations                                                                                                                                       
│   ├──exon_sums                                         
│   │   ├── human.exon_sums.G026.gtf.gz                                                                                                                    
│   └── gene_sums                                                                                                                                
│       ├── human.gene_sums.G026.gtf.gz                                                                                                                 
└── data_sources
    └── sratest
        ├── base_sums
        ├── exon_sums
        │   └── st
        │       └── sratest
        │           ├── sratest.exon_sums.sratest.ERCC.gz
        │           ├── sratest.exon_sums.sratest.F006.gz
        │           ├── sratest.exon_sums.sratest.G026.gz
        │           ├── sratest.exon_sums.sratest.G029.gz
        │           ├── sratest.exon_sums.sratest.R109.gz
        │           └── sratest.exon_sums.sratest.SIRV.gz
        ├── gene_sums                                                                                                                                    
        │   └── st                                                                                                                                                 
        │       └── sratest
        │           ├── sratest.gene_sums.sratest.ERCC.gz
        │           ├── sratest.gene_sums.sratest.F006.gz
        │           ├── sratest.gene_sums.sratest.G026.gz
        │           ├── sratest.gene_sums.sratest.G029.gz
        │           ├── sratest.gene_sums.sratest.R109.gz
        │           └── sratest.gene_sums.sratest.SIRV.gz
        ├── junctions
        │   └── st
        │       └── sratest
        │           ├── sratest.junctions.sratest.ALL.ID.gz
        │           ├── sratest.junctions.sratest.ALL.MM.gz
        │           ├── sratest.junctions.sratest.ALL.RR.gz
        │           ├── sratest.junctions.sratest.UNIQUE.ID.gz
        │           ├── sratest.junctions.sratest.UNIQUE.MM.gz
        │           └── sratest.junctions.sratest.UNIQUE.RR.gz
        └── metadata
            ├── sratest.recount_project.MD.gz
            └── st
                └── sratest
                    ├── sratest.recount_project.sratest.MD.gz
                    ├── sratest.recount_qc.sratest.MD.gz
                    └── sratest.sratest.sratest.MD.gz
```

Then we can load up recount3 and start working with this dataset:

```{r}
#| label: run-recount3
#| eval: false
library(recount3)
hp = available_projects(recount3_url = tmp_dir)
rse_genes = create_rse(hp[1, ], recount3_url = tmp_dir)
```