---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rc3helpers

<!-- badges: start -->
<!-- badges: end -->

rc3helpers is a little package designed around making it easier work with local samples in two ways:

1. Running recount-pump and recount-unify;
1. Copying the output of recount-unify into the directory structure that allows importing local data into the [recount3](https://bioconductor.org/packages/release/bioc/html/recount3.html) R package.


## Running Pump & Unify

This assumes you've already gotten everything setup for running pump and unify locally.
You can find instructions for that on the [monorail](https://github.com/langmeadlab/monorail) README.

I realize that the rc3_run_pump_unify function call takes a **lot** of arguments.
However, given that I'm horrible at writing shell scripts, and typically have more than a single sample to run on local hardware, not in the cloud
like the Langmead lab does, this saves me a lot of typing as I only have to give it the directory of samples, besides the other locations.

```{r}
#| label: running-pump-unify
#| eval: false
# NOT RUN, for example only
unify_output = rc3_run_pump_unify(
  fasta = "path/to/fasta.fq.gz",
  outputs = "path/to/outputs/dir",
  monorail = "path/to/monorail/repo",
  recount_pump = "path/to/recount-pump.XXX.sif",
  recount_unify = "path/to/recount-unify.XXX.sif",
  reference_path = "path/to/references",
  reference = "hg38",
  studyid = "studyid",
  shortid = "shortid",
  ncore = 2
)
```

### Custom Organisms

For the `monorail` argument, this can either be the top level directory that is normally *monorail-external*, or a vector of length 2, giving the paths to both
of the shell scripts for orchestrating *recount-pump* and *recount-unify* respectively.
This is useful when you are working with a custom organism outside of **human** and **mouse**.
For example, I have been working with **zebrafish**, so I created a slightly modified script that handles the *recount-unify* step properly for zebrafish, and
I would define the `monorail` argument like this, so that the unify step calls the correct shell script:

```r
monorail = c("monorail-external/singularity/run_recount_pump.sh", 
             "monorail-external/sngularity/run_recount_unify_mz11.sh")
```

### Samples and IDs

This script **currently** expects each sample to have two fasta files that are named:

* sample_id_1.fq.gz
* sample_id_2.fq.gz

*monorail-external* also assumes that the last two places of any identifier such as **sample_id**, **study_id** or **shortid** <span style="color: red">do not</span> contain an underscore or any other non-alphanumeric character, i.e. they should contain [a-z] and or [0-9]. 

## Copying Unify Outputs

Two important notes:

1. As of recount3 v 1.22.0, there is no option to have an organism outside of *human* or *mouse*, so we always use a directory of one of those two organisms.
1. This also means that only certain annotations / references are supported. Therefore, we create symbolic links to the files that `recount3` expects to be there.

## Installation

You can install the development version of `rc3helpers` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("MoseleyBioinformaticsLab/rc3helpers")
```

## Example

I previously ran two SRA samples in local mode through the pump and unify workflow, with the `short_id` **sratest**.

```{r}
#| label: setup-directories
library(rc3helpers)
tmp_dir = tempdir()
rc3_dirs = rc3_setup_directory(base_dir = tmp_dir, short_id = "sratest")
```

The directories created should look like this:

```
human
в”њв”Ђв”Ђ annotations
в”‚   в”њв”Ђв”Ђ exon_sums
в”‚   в””в”Ђв”Ђ gene_sums
в””в”Ђв”Ђ data_sources
    в””в”Ђв”Ђ sratest
        в”њв”Ђв”Ђ base_sums
        в”њв”Ђв”Ђ exon_sums
        в”њв”Ђв”Ђ gene_sums
        в”њв”Ђв”Ђ junctions
        в””в”Ђв”Ђ metadata
```

The unify output looks like this:

```
в”њв”Ђв”Ђ exon_sums_per_study
в”‚В В  в””в”Ђв”Ђ st
в”‚В В      в””в”Ђв”Ђ sratest
в”‚В В          в”њв”Ђв”Ђ sratest.exon_sums.sratest.ERCC.gz
в”‚В В          в”њв”Ђв”Ђ sratest.exon_sums.sratest.F006.gz
в”‚В В          в”њв”Ђв”Ђ sratest.exon_sums.sratest.G026.gz
в”‚В В          в”њв”Ђв”Ђ sratest.exon_sums.sratest.G029.gz
в”‚В В          в”њв”Ђв”Ђ sratest.exon_sums.sratest.R109.gz
в”‚В В          в””в”Ђв”Ђ sratest.exon_sums.sratest.SIRV.gz
в”њв”Ђв”Ђ gene_sums_per_study
в”‚В В  в””в”Ђв”Ђ st
в”‚В В      в””в”Ђв”Ђ sratest
в”‚В В          в”њв”Ђв”Ђ sratest.gene_sums.sratest.ERCC.gz
в”‚В В          в”њв”Ђв”Ђ sratest.gene_sums.sratest.F006.gz
в”‚В В          в”њв”Ђв”Ђ sratest.gene_sums.sratest.G026.gz
в”‚В В          в”њв”Ђв”Ђ sratest.gene_sums.sratest.G029.gz
в”‚В В          в”њв”Ђв”Ђ sratest.gene_sums.sratest.R109.gz
в”‚В В          в””в”Ђв”Ђ sratest.gene_sums.sratest.SIRV.gz
в”њв”Ђв”Ђ ids.tsv
в”њв”Ђв”Ђ junction_counts_per_study
в”‚В В  в””в”Ђв”Ђ st
в”‚В В      в””в”Ђв”Ђ sratest
в”‚В В          в”њв”Ђв”Ђ sratest.junctions.sratest.ALL.ID.gz
в”‚В В          в”њв”Ђв”Ђ sratest.junctions.sratest.ALL.MM.gz
в”‚В В          в”њв”Ђв”Ђ sratest.junctions.sratest.ALL.RR.gz
в”‚В В          в”њв”Ђв”Ђ sratest.junctions.sratest.UNIQUE.ID.gz
в”‚В В          в”њв”Ђв”Ђ sratest.junctions.sratest.UNIQUE.MM.gz
в”‚В В          в””в”Ђв”Ђ sratest.junctions.sratest.UNIQUE.RR.gz
в”њв”Ђв”Ђ junctions.bgz
в”њв”Ђв”Ђ junctions.bgz.tbi
в”њв”Ђв”Ђ junctions.sqlite
|
|
|
```

Where this is a model organism, you are going to need the reference sets of things downloaded too.
We can download the gene and exon sum annotations from recount3, in our case for G026.

```
wget http://duffel.rail.bio/recount3/human/new_annotations/gene_sums/human.gene_sums.G026.gtf.gz
wget http://duffel.rail.bio/recount3/human/new_annotations/exon_sums/human.exon_sums.G026.gtf.gz
```

We can pass the list of directories created by `rc3_setup_directories` to `rc3_copy_unify_outputs`:

```{r}
#| label: copy-unify
#| eval: false
rc3_dirs |>
  rc3_copy_unify_outputs(
    short_id = "sratest",
    unify_directory = "/path/to/unify",
    references_directory = "/path/to/references/hg38"
  )
```

Which will create this directory structure. You may see an error when working with human data, as the function tries to link a file that actually already exists.
That's OK.

```
human
в”њв”Ђв”Ђ annotations                                                                                                                                       
в”‚В В  в”њв”Ђв”Ђexon_sums                                         
в”‚В В  в”‚В В  в”њв”Ђв”Ђ human.exon_sums.G026.gtf.gz                                                                                                                    
в”‚В В  в””в”Ђв”Ђ gene_sums                                                                                                                                
в”‚В В      в”њв”Ђв”Ђ human.gene_sums.G026.gtf.gz                                                                                                                 
в””в”Ђв”Ђ data_sources
    в””в”Ђв”Ђ sratest
        в”њв”Ђв”Ђ base_sums
        в”њв”Ђв”Ђ exon_sums
        в”‚В В  в””в”Ђв”Ђ st
        в”‚В В      в””в”Ђв”Ђ sratest
        в”‚В В          в”њв”Ђв”Ђ sratest.exon_sums.sratest.ERCC.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.exon_sums.sratest.F006.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.exon_sums.sratest.G026.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.exon_sums.sratest.G029.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.exon_sums.sratest.R109.gz
        в”‚В В          в””в”Ђв”Ђ sratest.exon_sums.sratest.SIRV.gz
        в”њв”Ђв”Ђ gene_sums                                                                                                                                    
        в”‚В В  в””в”Ђв”Ђ st                                                                                                                                                 
        в”‚В В      в””в”Ђв”Ђ sratest
        в”‚В В          в”њв”Ђв”Ђ sratest.gene_sums.sratest.ERCC.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.gene_sums.sratest.F006.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.gene_sums.sratest.G026.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.gene_sums.sratest.G029.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.gene_sums.sratest.R109.gz
        в”‚В В          в””в”Ђв”Ђ sratest.gene_sums.sratest.SIRV.gz
        в”њв”Ђв”Ђ junctions
        в”‚В В  в””в”Ђв”Ђ st
        в”‚В В      в””в”Ђв”Ђ sratest
        в”‚В В          в”њв”Ђв”Ђ sratest.junctions.sratest.ALL.ID.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.junctions.sratest.ALL.MM.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.junctions.sratest.ALL.RR.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.junctions.sratest.UNIQUE.ID.gz
        в”‚В В          в”њв”Ђв”Ђ sratest.junctions.sratest.UNIQUE.MM.gz
        в”‚В В          в””в”Ђв”Ђ sratest.junctions.sratest.UNIQUE.RR.gz
        в””в”Ђв”Ђ metadata
            в”њв”Ђв”Ђ sratest.recount_project.MD.gz
            в””в”Ђв”Ђ st
                в””в”Ђв”Ђ sratest
                    в”њв”Ђв”Ђ sratest.recount_project.sratest.MD.gz
                    в”њв”Ђв”Ђ sratest.recount_qc.sratest.MD.gz
                    в””в”Ђв”Ђ sratest.sratest.sratest.MD.gz
```

Then we can load up recount3 and start working with this dataset:

```{r}
#| label: run-recount3
#| eval: false
library(recount3)
hp = available_projects(recount3_url = tmp_dir)
rse_genes = create_rse(hp[1, ], recount3_url = tmp_dir)
```